---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png",
  dpi = 300
)
```

# `r emo::ji("fireworks")` fireworking

Playing around with data from U.S. Consumer Products Safety Commission (CPSC) [2020 Fireworks Annual Report](https://www.cpsc.gov/s3fs-public/2020-Fireworks-Annual-Report.pdf).[^1]

```{r libs, message=FALSE}
library(tidyverse)
library(tabulizer)
library(pdftools)
# plus some bonus appearances from other packages,
# e.g. {janitor}, {hrbrthemes}, and {scales}
```
The report is available online as a pdf, so let's tell R about that.

```{r pdf-url}
pdf_file <- "https://www.cpsc.gov/s3fs-public/2020-Fireworks-Annual-Report.pdf"
```

Now we'll extract the table on p.14 of the report using [{tabulizer}](https://github.com/ropensci/tabulizer), and do a little cleaning up with an assist from [{janitor}](https://sfirke.github.io/janitor/index.html).

```{r table-one}
p14_tables <- extract_tables(pdf_file, pages = 14, output = "data.frame")

# returns a list, so take first item from list
injuries_by_year <- p14_tables[[1]]

injuries_by_year <- injuries_by_year %>%
  as_tibble() %>%
  janitor::clean_names() %>%
  mutate(estimated_injuries = as.numeric(str_remove(estimated_injuries, ","))) %>%
  rename("injuries_per_100000" = "injuries_per_100_000_people")

injuries_by_year
```
Now let's plot it to see the general trend visually.

```{r injuries-by-year, fig.alt="Estimated fireworks-related injuries from 2005 to 2020: A scatterplot showing an increasing trend in estimated-injuries from 2008 to 2020"}
ggplot(injuries_by_year, aes(x = year, y = estimated_injuries)) +
  geom_point() +
  scale_y_continuous(label = scales::label_comma()) +
  labs(title = "Estimated fireworks-related injuries 2005-2020",
       caption = "source: NEISS, U.S. CPSC",
       alt = paste("A scatterplot showing an increasing trend in",
                   "estimated-injuries from 2008 to 2020")) +
  ylab("estimated injuries") +
  hrbrthemes::theme_ipsum()
```

Looks like 2020 was a big year for getting injured with fireworks (which is probably a surprise to no one).

Let's see if we can get the data from one of the more complicated (read: multiple-headered, non-tidy) tables, like the one on page 24 breaking down 2020 injuries by body region and type.

```{r p24-table}
p24_tables <- extract_tables(pdf_file, pages = 24, output = "data.frame")

inj_by_body_region <- tibble(p24_tables[[1]]) # tibble for the printing

inj_by_body_region
```
OK, it's not the worst thing in the world, but it needs some fixing up. We've got the headers we _actually_ want in the first row, and we've got two sets of values in the row currently labelled "Diagnosis". I'm sure the latter problem could be dealt with in a nicer way using regular expressions, or something of the sort, but I'm gonna cheat and just replace it to make my life easier.


```{r}
inj_by_body_region[1,4] <- "Contusions/Lacerations Fractures/Sprains"
```

Since [{janitor}](https://sfirke.github.io/janitor) has a bunch of functions that are handy for cleaning things like this up, let's go ahead and just load the library. I'm also going to use `separate()` on the empty (-looking) rows to my advantage, so I can filter them out later with `is.na()`.

```{r, message=FALSE}
library(janitor)

inj_by_body_region %>%
  separate(Diagnosis, c("diagnosis1", "diagnosis2"), sep = " ") %>%
  filter(!is.na(diagnosis2)) %>%
  row_to_names(1) %>%
  clean_names() -> inj_body_region

inj_body_region
```

Looking pretty good! Even though the `*` denotes "estimates of fewer than 50 injuries," I'm going to replace them with zeroes so that we can treat the column (and all the other number-y ones) as numeric. We'll also ditch the row with "Total" listed as a `body_region`, since it's not one, as well as the `total` column, since we know how to add.

```{r}
inj_body_region %>%
  mutate("fractures_sprains" = replace(fractures_sprains, fractures_sprains == "*", "0")) %>%
  mutate(across(-body_region, ~ as.numeric(str_remove(.x, ",")))) %>%
  filter(body_region != "Total") %>%
  select(-total) -> inj_body_region2

inj_body_region2
```

Our data _still_ isn't quite in the tidy form I want (though it does make for a nice, concise-looking table). Let's `pivot_longer()` to make the type of injury a variable/column.

```{r}
tidy_inj_body_region <- inj_body_region2 %>%
  pivot_longer(!body_region, names_to = "injury", values_to = "count")

tidy_inj_body_region
```
At this point, do I regret making those clean column names? Maybe a little…but I really hate dealing with backticks. Plus, I could always go back in and make things pretty now.

---

[^1]: Marier, A., Smith, B., Lee, S. (2021, June). _2020 Fireworks Annual Report: Fireworks-related deaths, emergency depertment-treated injuries, and enforcement activities during 2020_. U.S. Consumer Product Safety Commission. https://www.cpsc.gov/s3fs-public/2020-Fireworks-Annual-Report.pdf
